Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-01-31 paper/main.tex
Updates

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations}
We consider random variables $ X $ and $ Y $ with a possible multi-dimensional
variable $ \bm{Z} = \{ Z_1, \cdots, Z_k \} $. We consider these variables in
the mixed data setting where these variables can be any combination of
continuous, categorical, or ordinal unless specified. We also assume all
variables to be observed. We use $ \rvert \bm{Z} \rvert $ to represent the
cardinality of the variable $ \bm{Z} $. We write $ \bm{x} = (x_1, \cdots, x_n)
$ for a sample from $ X $ of size $ n $. We write the expectation of a variable
$ X $ as $ \mathbb{E}[X] $, conditional expectations as $ \mathbb{E}[X |
\bm{Z}] $.

We consider the problem of estimating the partial (or conditional) association
between $ X $ and $ Y $ in the presence of a set of conditioning variables $
\bm{Z} $. When $ Z = \emptyset $, this is equivalent to the marginal
association between $ X $ and $ Y $.

\subsection{Measures of Conditional Association}
As our DAG construction procedure relies on a measure of conditional
association between variables, in this section we list out some of the already
known measures for specific data types. These measure of associations are
typically the effect size of various conditional independence tests but they
don't need to be. However, there are no generalized measure of association for
mixed data, that we propose in the next section. This is by no means an
exhaustive list, rather just to give an idea of what can be done.

\paragraph{Both $ X $ and $ Y $ are continuous}
Where there are no conditional variables, $ \bm{Z} = \emptyset $, Pearson's
correlation coefficient can be used as a measure of association between the
variables. This does make various assumptions such as linearity between the
variables, normal distribution for $ X $ and $ Y $, and homoscedasticity.

In the case when $ \bm{Z} \neq \emptyset $, then partial correlations can be
used. This requires using two linear regression models: $ E_X: X \sim \bm{Z} $
and $ E_Y: Y \sim \bm{Z} $. Then computing the residuals using these two
regressions as: $ R_X = X - E_X(\bm{Z}) $ and $ R_Y = Y - E_Y(\bm{Y}) $. After
this a simple correlation coefficient can be computed between $ R_X $ and $ R_Y $.

@@

\section{A Measure of Association for Mixed Data}
\label{sec:mixed_association}

In this section, we introduce a novel measure of partial association for mixed
data by extending the concept of partial correlations \citep{Whittaker2009}
\todo{Verify this citation} to handle mixed data types. To compute partial
correlations, we have two key steps: 1) a residualization method, and 2) a
measure of associations between the residuals. When both $ X $ and $ Y $ are
continuous variables, residuals are typically computed using a regression-based
method, and Pearson's correlation coefficient is used to measure the association
between these residuals. To extend this approach to mixed data, we combine a mixed
data residualization method \citep{Ankan2023} with a multivariate measure of
association based on canonical correlations, Pillai's Trace \citep{Pillai1955}.

Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y $, and $
\bm{Z} $, we aim to estimate the conditional
association $ \rho(X, Y; \bm{Z}) $ between $ X $ and $ Y $ conditioned on $
\bm{Z} $. We use the residualization approach from \citet{Ankan2023} to compute 
the residuals. The residuals are defined as follows, depending on the type of 
dependent variable:

@@

	$ G $ <- Empty Graph
	p_values, effects <- compute_effects(G)

\end{verbatim}

>>>>

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be continuous, categorical, or ordinal unless specified.

We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $, conditional
expectations as $ \mathbb{E}[X | \bm{Z}] $.

\subsection{Measures of Conditional Association}
Our proposed method for iterative DAG construction relies on a measure of
conditional association between variables. We are interested in measuring the
association between variables $ X $ and $ Y $ when conditioned on a set of
variables $ \bm{Z} $ which is potentially an empty set, $ \bm{Z} = \emptyset $.
Depending on the type of $ X $, $ Y $, and $ \bm{Z} $ various measures of
associations have been used. Typically, these are the effect size measures of
the conditional independence test $ X \ci Y \rvert \bm{Z} $. Here, we give an
overview of some of the commonly used measures for different data types.

\subsubsection{Both $ X $ and $ Y $ are continuous}
In the case when both $ X $ and $ Y $ are continuous variables with the
assumptions of normality and linearly relationship, typically Pearson's
correction coefficients are used. In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient can be directly computed. When $ \bm{Z} \neq \emptyset
$, partial correlations are used. One of the ways to estimate it is to use two
linear regression models: $ E_X: x \sim \bm{z} $, and $ E_Y: y \sim \bm{z} $.
Then we take the compute the Pearson's correlation coefficient between the
residuals of these two regressions: $ R_X = x - E_X(\bm{z}) $, $ R_Y = y -
E_Y(\bm{z}) $.

@@

\section{A Generalized Measure of Conditional Association}
\label{sec:mixed_association}

As there are no generalized measures of association for mixed data, in this
section, we first introduce a novel measure of partial association for mixed
data by extending the concept of partial correlations \citep{Whittaker2009}
\todo{Verify this citation} to handle mixed data types. To compute partial
correlations, we have two key steps: 1) a residualization method, and 2) a
measure of associations between the residuals. When both $ X $ and $ Y $ are
continuous variables, residuals are typically computed using a regression-based
method, and Pearson's correlation coefficient is used to measure the
association between these residuals. To extend this approach to mixed data, we
combine a mixed data residualization method \citep{Ankan2023} with a
multivariate measure of association based on canonical correlations, Pillai's
Trace \citep{Pillai1955}.

Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y $, and $
\bm{Z} $, we aim to estimate the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We use the residualization
approach from \citet{Ankan2023} to compute the residuals. For each of the
variable $ X $ and $ Y $, we compute the residual $ R_X $ and $R_Y$
respectively as follows depending on their type.

@@

	$ G $ <- Empty Graph
	p_values, effects <- compute_effects(G)
	while (p_values > p_thres, 
\end{verbatim}
